Bayesian model comparison for rare variant association studies of multiple phenotypes

### 2018-01-30 tex/manuscript.tex
Manuscript updates

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Whole genome sequencing studies applied to large populations or biobanks with extensive phenotyping raise new analytic challenges. The need to consider many variants at a locus or group of genes simultaneously and the potential to study many correlated phenotypes with shared genetic architecture provide opportunities for discovery and inference that are not addressed by the traditional one variant-one phenotype association study. Here we introduce a model comparison approach we refer to as MRP for rare variant association studies that considers correlation, scale, and location of genetic effects across a group of genetic variants, phenotypes, and studies. We consider the use of summary statistic data to apply 1) standard univariate and multivariate gene-based meta-analysis models, 2) models for assessing heterogeneity of genetic effects, and 3) models for identifying protective protein-truncating variants that can expedite drug discovery.  Through simulation studies, we demonstrate that the proposed model comparison approach may improve ability to detect rare variant association signals. Finally, we demonstrate the model on two groups of phenotypes from the UK Biobank: 1) asthma diagnosis, eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor. We demonstrate that the model leverages information from multiple phenotypes to identify known and novel rare variant associations for asthma and glaucoma. Overall the model comparison approach is able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.

@@

%%Comparing to other methods 
To evaluate the performance of MRP and to study its behavior we use simulations and compare it to other commonly used approaches. Some simple alternatives to MRP include univariate approaches for rare variant association studies including the sequence kernel association test (SKAT)\cite{skat}, and the burden test, which we show are special cases of the MRP model comparison when we assign the prior correlation of genetic effects across different variants to be zero or one. Furthermore, MRP can be extended to detect the presence of heterogeneity of effects, which in some circumstances, such as those where effects are observed to be heterogeneous across studies, may indicate the presence of technical sources of error.

@@

We also found moderate evidence for associations between rare coding variants in \textit{CCR3} and asthma. The $\log_{10}$ BFs for \textit{CCR3} were 3.3 and 3.1 in the joint and joint protective analyses respectively compared to only -0.5 in the asthma-only analysis  (Fig~\ref{asthma_bma}, Table~\ref{asthma_table}). \textit{CCR3} was not reported in a large GWAS for allergic disease including asthma\cite{Ferreira:2017ba} though \textit{CCR3} is near a locus associated with atopy in a previous meta-analysis\cite{Ober:2011jk}. Other genes like \textit{IL18RAP}, \textit{ATP2A3}, and \textit{FLG} had $\log_{10}$ BFs greater than 4 in the asthma-only analysis but much smaller BFs in the joint analyses indicating that rare variants in these genes are less likely to affect this group of traits. Similarly, there were other genes like \textit{RP11-39K24.9} and \textit{IL17RA} that had relatively larger BFs in the eosinophil count-only analysis but small BFs for the joint analyses demonstrating MRP's ability to integrate information across all phenotypes considered.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Whole genome sequencing studies applied to large populations or biobanks with extensive phenotyping raise new analytic challenges. The need to consider many variants at a locus or group of genes simultaneously and the potential to study many correlated phenotypes with shared genetic architecture provide opportunities for discovery and inference that are not addressed by the traditional one variant-one phenotype association study. Here we introduce a model comparison approach we refer to as MRP for rare variant association studies that considers correlation, scale, and location of genetic effects across a group of genetic variants, phenotypes, and studies. We consider the use of summary statistic data to apply univariate and multivariate gene-based meta-analysis models for identifying rare variant associations with an emphasis on protective protein-truncating variants that can expedite drug discovery. Through simulation studies, we demonstrate that the proposed model comparison approach can improve ability to detect rare variant association signals. Finally, we apply the model to two groups of phenotypes from the UK Biobank: 1) asthma diagnosis, eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor. We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma. We also find evidence for an association between rare variants in the drug target \textit{CCR3} and asthma. We identified novel protective associations between rare variants in \textit{ANGPTL7} and glaucoma as well. Overall, we show that the model comparison approach is able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.

@@

%%Comparing to other methods 
To evaluate the performance of MRP and to study its behavior we use simulations and compare it to other commonly used approaches. Some simple alternatives to MRP include univariate approaches for rare variant association studies including the sequence kernel association test (SKAT)\cite{skat}, and the burden test, which we show are special cases of the MRP model comparison when we assign the prior correlation of genetic effects across different variants to be zero or one.

@@

We also found moderate evidence for a protective association between rare coding variants in \textit{CCR3} and asthma. The $\log_{10}$ BFs for \textit{CCR3} were 3.3 and 3.1 in the joint and joint protective analyses respectively compared to only -0.5 in the asthma-only analysis  (Fig~\ref{asthma_bma}, Table~\ref{asthma_table}). \textit{CCR3} is a chemokine receptor that is highly expressed on eosinophils and has been a therapeutic focus for asthma \cite{Neighbour:2014jw, Pease:2014em}. \textit{CCR3} was not reported in a large GWAS for allergic disease including asthma\cite{Ferreira:2017ba} though \textit{CCR3} is near a locus associated with atopy in a previous meta-analysis\cite{Ober:2011jk}. These results demonstrate that MRP can identify biologically meaningful therapeutic targets that may be missed by standard GWAS approaches.

Considering multiple phenotypes jointly allows for the efficient prioritization of disease genes. For instance, some genes like \textit{IL18RAP}, \textit{ATP2A3}, and \textit{FLG} had $\log_{10}$ BFs greater than 4 in the asthma-only analysis but much smaller BFs in the joint analyses indicating that rare variants in these genes are less likely to affect this group of traits. Similarly, there were other genes like \textit{RP11-39K24.9} and \textit{IL17RA} that had relatively larger BFs in the eosinophil count-only analysis but small BFs for the joint analyses demonstrating MRP's ability to integrate information across all phenotypes considered.
